Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-04-02 optim.qmd
done with section 2

@cook1995grand introduced the PP guided tour, which enabled interactive visualisation of the optimisation in order to visually explore high-dimensional data. It is implemented in the R [XXX REF] package `tourr` [@tourr]. The optimisation that is implemented is fairly basic, and potential problems were highlighted by @RJ-2021-105. Implementing better optimisation functionality is a goal, but it needs to be kept in mind that the guided tour also has places importance on watching the projected data as the optimisation progresses.

@@

>>>>

@cook1995grand introduced the PP guided tour, which enabled interactive visualisation of the optimisation in order to visually explore high-dimensional data. It is implemented in the R [@R] package `tourr` [@tourr]. The optimisation that is implemented is fairly basic, and potential problems were highlighted by @RJ-2021-105. Implementing better optimisation functionality is a goal, but it needs to be kept in mind that the guided tour also has places importance on watching the projected data as the optimisation progresses.

@@

A tour on high-dimensional data is constructed by geodesically interpolating between pairs of planes. Any plane is described by an orthonormal basis, $A_t$, where $t$ represents time in the sequence. The term "geodesic" refers to maintaining the orthonormality constraint so that each view shown is correctly a projection of the data. The PP guided tour operates by geodesically interpolating to target planes (projections) which have high PP index values, as provided by the optimiser. The geodesic interpolation means that the viewer sees a continuous sequence of projections of the data, so they can watch patterns of interest forming as the function is optimised. There are five optimisation methods implemented in the `tourr` package:

- `search_geodesic()`: provides a pseudo-derivative optimisation. It searches locally for the best direction, based on differencing the index values for very close projections. Then it follows the direction along the geodesic path between planes, stopping when the next index value fails to increase. 
- `search_better()`: is a brute-force optimisation searching randomly for projections with higher index values. 
- `search_better_random()`: is essentially simulated annealing [@Bertsimas93] where the search space is reduced as the optimisation progresses. 
- `search_posse()`: implements the algorithm described in @posse95. 
- `search_polish()`: is a very localised search, to take tiny steps to get closer to the local maximum.

There are several PP index functions available: `holes()` and `cmass()` [@cook1993projection]; `lda_pp()` [@lee2005projection]; `pda_pp()` [@lee2010projection]; `dcor2d()` and `splines2d()` [@Grimm2016];  `norm_bin()` and `norm_kol()` [@huber85]; `slice_index()` [@Laa:2020wkm]. Most are relatively simply defined, for any projection dimension, and implemented because they are relatively easy to optimise. A goal is to be able to incorporate more complex PP indexes, for example based on scagnostics (@scag, @WW08).

An initial investigation of PP indexes, and the potential for scagnostics is described in @laa_using_2020. To be useful here an optimiser needs to be able to handle functions which are not very smooth. In addition, because data structures might be relatively fine, the optimiser needs to be able to find maxima that occur with a small squint angle, that can only be seen from very close by. One last aspect that is useful is for an optimiser to return local maxima in addition to global because data can contain many different and interesting features.

### 2024-04-02 optim.tex
done with section 2

\citet{cook1995grand} introduced the PP guided tour, which enabled
interactive visualisation of the optimisation in order to visually
explore high-dimensional data. It is implemented in the R {[}XXX REF{]}
package \texttt{tourr} \citep{tourr}. The optimisation that is
implemented is fairly basic, and potential problems were highlighted by
\citet{RJ-2021-105}. Implementing better optimisation functionality is a
goal, but it needs to be kept in mind that the guided tour also has
places importance on watching the projected data as the optimisation
progresses.

@@

>>>>

\citet{cook1995grand} introduced the PP guided tour, which enabled
interactive visualisation of the optimisation in order to visually
explore high-dimensional data. It is implemented in the R \citep{R}
package \texttt{tourr} \citep{tourr}. The optimisation that is
implemented is fairly basic, and potential problems were highlighted by
\citet{RJ-2021-105}. Implementing better optimisation functionality is a
goal, but it needs to be kept in mind that the guided tour also has
places importance on watching the projected data as the optimisation
progresses.

@@

A tour on high-dimensional data is constructed by geodesically
interpolating between pairs of planes. Any plane is described by an
orthonormal basis, \(A_t\), where \(t\) represents time in the sequence.
The term ``geodesic'' refers to maintaining the orthonormality
constraint so that each view shown is correctly a projection of the
data. The PP guided tour operates by geodesically interpolating to
target planes (projections) which have high PP index values, as provided
by the optimiser. The geodesic interpolation means that the viewer sees
a continuous sequence of projections of the data, so they can watch
patterns of interest forming as the function is optimised. There are
five optimisation methods implemented in the \texttt{tourr} package:

\begin{itemize}
\tightlist
\item
  \texttt{search\_geodesic()}: provides a pseudo-derivative
  optimisation. It searches locally for the best direction, based on
  differencing the index values for very close projections. Then it
  follows the direction along the geodesic path between planes, stopping
  when the next index value fails to increase.
\item
  \texttt{search\_better()}: is a brute-force optimisation searching
  randomly for projections with higher index values.
\item
  \texttt{search\_better\_random()}: is essentially simulated annealing
  \citep{Bertsimas93} where the search space is reduced as the
  optimisation progresses.
\item
  \texttt{search\_posse()}: implements the algorithm described in
  \citet{posse95}.
\item
  \texttt{search\_polish()}: is a very localised search, to take tiny
  steps to get closer to the local maximum.
\end{itemize}

There are several PP index functions available: \texttt{holes()} and
\texttt{cmass()} \citep{cook1993projection}; \texttt{lda\_pp()}
\citep{lee2005projection}; \texttt{pda\_pp()} \citep{lee2010projection};
\texttt{dcor2d()} and \texttt{splines2d()} \citep{Grimm2016};
\texttt{norm\_bin()} and \texttt{norm\_kol()} \citep{huber85};
\texttt{slice\_index()} \citep{Laa:2020wkm}. Most are relatively simply
defined, for any projection dimension, and implemented because they are
relatively easy to optimise. A goal is to be able to incorporate more
complex PP indexes, for example based on scagnostics (\citet{scag},
\citet{WW08}).

An initial investigation of PP indexes, and the potential for
scagnostics is described in \citet{laa_using_2020}. To be useful here an
optimiser needs to be able to handle functions which are not very
smooth. In addition, because data structures might be relatively fine,
the optimiser needs to be able to find maxima that occur with a small
squint angle, that can only be seen from very close by. One last aspect
that is useful is for an optimiser to return local maxima in addition to
global because data can contain many different and interesting features.
